An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-10-02 paper.tex
Update §5.3.2

The communication patterns that different levels in the hierarchy
experience may be  indicative of the cooperation between
levels~\cite{li2021does,duan2018authoritarian}. For instance, if
participants at higher levels send lots of email while receiving little,
this suggests a lack of receptiveness of higher levels to the suggestions
of lower levels, and low confidence of lower levels to voice their
opinions. In contrast, we hypothesise that collaboration between layers is
high, as outlined in the IETF mission statement, which will look more like
the opposite of this example.

In this vein, we compare the ways that different roles communicate, looking
at the tendency for communications to be purely inbound, outbound or use
more complex patterns. We count three edge motifs for different hierarchy
levels as described in section~\ref{sec:motif}. We find that higher levels
consistently experience a higher proportion of incoming communication than
RPs, and all levels send outward communication at similar proportions over
time. Also, the proportion of people who begin email threads are split by
hierarchy level. Higher levels show a disproportionate tendency to
originate email threads.

@@

To quantify the effect of being in each hierarchy level has on network wide
communication for \textbf{RQ2}, we calculate the proportion of three node
motifs, lasting at most a month, in which each node participates over a
year period, see section~\ref{sec:motif}. These motifs are counted,
categorised into the three hierarchy levels and then proportions are taken
of each motif type.

Figure~\ref{fig:motif_proportions} shows the proportions of each level for
three node ``Inward Star", ``Outward Star", ``Mixed Star", and
``Triangles". The Inward and Mixed Star motifs show about a $4\%$ and
$10\%$ increase in proportion for higher levels in the hierarchy versus for
RPs. However, about a $10\%$ increase for RPs is seen for Triangles versus
WGCs and ADs, and Outward Star proportions remain similar for all levels of
the hierarchy.

The larger proportion for WGCs and ADs of Inward and Mixed Star motifs
whilst the Outward Star motifs remain similar for all levels suggests that
higher levels receive more direct communication. The higher proportion of
RP Triangle motifs suggests their discussion is more of a group activity,
whereas WGCs and ADs have a larger proportion of one-on-one conversations.
Therefore, we interpret the general communication patterns within the IETF
as a discussion amongst RPs interspersed with questions sent to and
announcements from WGCs and ADs. This indicates some collaboration between
layers and confidence of some RPs' to voice their opinions upwards.

@@

Figure~\ref{fig:origin_emails} seems to corroborate this. We calculate the
proportion of mailing list threads which nodes in each hierarchy level
originate, using a year window, pushed forward by a month each calculation.
We see that WGCs and ADs send a disproportionate number of originating
emails to the mailing lists. WGCs send 30--50\% whereas they are 6--10\% of
active individuals, and ADs 5--10\% versus 0.5\%.

Therefore, the larger proportion of inward motifs for higher levels is in
part due to WGCs and ADs disproportionately originating email threads.
Other IETF participants will then reply to the thread, boosting the
proportion of inward motifs WGCs and ADs appear in. This may suggest the
higher levels are good ``condensers of communication" as they receive more
than they send.

In answer to the RQ, combining the interpretations of the two plots, the
discussion seems to happen in the following way: The ADs or WGCs are
commonly the originators of discussion threads. Where conversation includes
a WGC or AD the conversation tends to be directed around them and usually
inward towards the WGC or AD. The RPs are more likely to engage in
discussions amongst themselves (triangular communication patterns) when
higher level individuals are not present.

@@

>>>>

In this section, we explore communication patterns within the IETF, to
understand how participants in different roles (regular participants,
working group chairs, area directors) communicate and whether those
patterns of communication are indicative of healthy cooperation in the
running of the organisation \cite{li2021does,duan2018authoritarian}.

For instance, if those in leadership positions send lots of email while
receiving little, this suggests a lack of receptiveness of the leadership
to the suggestions from regular participants, and low confidence of those
regular participants to voice their opinions. In contrast, we hypothesise
that if the IETF operates as outlined in its mission statement, in a more
cooperative manner, then communication between those in different roles
will be more symmetric.

\pb{Temporal Motifs:}
We compare the ways that people in different roles communicate, looking
at the tendency for communications to be purely inbound, purely outbound,
or to follow more complex patterns. 
%
% From our ICWSM 2024 paper, Section 2:
%
We describe these in terms of \emph{temporal motifs}: patterns in the
sequence in which edges are added to the communication graph over time.
Specifically, we analyse every possible combination of three temporally
ordered links~\cite{paranjape2017motifs} between at most three nodes.
The time order in which the nodes occur is important.  The direction of
edge 1 is arbitrary, but every edge temporally after it has their direction
oriented with respect to it.

Figure \ref{fig:motif_types} shows all possible types of interaction and
classifies them into five types: motifs with two nodes; ``Outward Star",
announcements or dissemination of information; ``Inward Star", questions or
condensing of information; ``Mixed Star", one-on-one discussion; and
``Triangle" is group discussion with no individual as the focus. Motifs
with only two nodes  are rare in this network and we do not present the
results.

\begin{figure}
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/motifs.pdf}
  \caption{
    Classification of three edge motif types with the numbers representing
    the order of communication.  Light blue squares are  ``Outward Star''
    motifs; purple are ``Inward Star'' motifs; orange are ``Mixed Star''
    motifs and grey are ``Triangle'' motifs. The box in the bottom left hand
    corner represents motifs with only two nodes.
  }
  \label{fig:motif_types}
\end{figure}

Raphtory~\cite{steer2020raphtory} is used to count the prevalence of these
motifs. We start with the list of 36 different combinations of three
directed edges shown in figure~\ref{fig:motif_types}. The graph is then
split into time windows and then the motifs are counted combinatorially for
each window; each of a node's edges are followed from their origin to three
edge steps away. Finally, we combine the motif counts using the categories
shown in figure~\ref{fig:motif_types} by the colour of the squares.

% We find that those in leadership positions (working group chairs, area
% directors) consistently experience a higher proportion of incoming
% communication than regular participants, and all levels send outward
% communication at similar proportions over time. Also, the proportion of
% people who begin email threads are split by hierarchy level. Higher levels
% show a disproportionate tendency to originate email threads.

@@

\pb{Impact of Role on Communication Patterns:}
To quantify the effect of being in a leadership position has on network
wide communication, we calculate the proportion of three node temporal
motifs, lasting at most a month, in which each node participates over a
year period. These temporal motifs are counted, categorised into the three
hierarchy levels (regular participant, working group chair, area director)
and then proportions are taken of each temporal motif type.

Figure~\ref{fig:motif_proportions} shows the proportions of each level for
three node ``Inward Star", ``Outward Star", ``Mixed Star", and
``Triangles". The Inward and Mixed Star motifs show about a $4\%$ and
$10\%$ increase in proportion for higher levels in the hierarchy versus
regular participants. However, about a $10\%$ increase for regular
participants is seen for Triangles versus working group chairs and area
directors, and Outward Star proportions remain similar for all levels of
the hierarchy.

The larger proportion of those in leadership roles in Inward and Mixed Star
temporal motifs, whilst Outward Star motifs remain similar for all levels,
suggests that those in leadership receive more direct communication. The
higher proportion of Triangle motifs for regular participants suggests
their discussion is more of a group activity, whereas working group chairs
and area directors have a larger proportion of one-on-one conversations.  
We therefore interpret the general communication patterns within the IETF
as a discussion amongst participants interspersed with questions sent to,
and announcements from, the working group chairs and area directors.  This
indicates some collaboration between layers and confidence of some regular
participants' to voice their opinions upwards.

@@

Figure~\ref{fig:origin_emails} seems to corroborate this. We calculate the
proportion of mailing list threads which nodes in each hierarchy level
originate, using a year window, pushed forward by a month each calculation.
We see that working group chairs and area directors send a disproportionate
number of originating emails to the mailing lists (working group chairs
send 30--50\% whereas they are 6--10\% of active individuals, and area
directors 5--10\% versus 0.5\%). 
Therefore, the larger proportion of inward motifs for those in leadership
roles is in part due to their disproportionately originating email threads.
Other IETF participants will then reply to the thread, boosting the
proportion of inward motifs working group chairs and area directors appear
in. This may suggest the higher levels are good ``condensers of
communication" as they receive more than they send.

@@

\pb{Summary:}
Combining the interpretations of Figures \ref{fig:motif_proportions} and
\ref{fig:origin_emails}, the discussion seems to happen in the following
way: area directors or working group chairs are commonly the originators
of discussion threads. Where conversation includes a working group chair
or area director the conversation tends to be directed around them and
usually inward towards the working group chair or area director. The
regular participants are more likely to engage in discussions amongst
themselves (triangular communication patterns) when higher level
individuals are not present.
